Adversarial Variational Optimization\\ of Non-Differentiable Simulators

### 2018-02-11 colt/avo.tex
keywords
save

\begin{keywords}
List of keywords
\end{keywords}

@@

As explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} (WGANs) reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, such that
\begin{align}
& W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) \nonumber \\
&= \sup_{\bfphi}  \mathbb{E}_{\tilde{\mathbf{x}}\sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}}; \bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x}; \bfphi)] \nonumber \\
&\triangleq \sup_{\bfphi} \mathcal{L}_W.
\end{align}
Under the WGAN-GP formulation of \citep{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\bfphi$ and $\bftheta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,&  \mathcal{L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - \mathcal{L}_W
\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

The observed data is sampled from a Poisson with mean $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 3000 iterations with mini-batches of size $M=50$
and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=100$ iterations with a step size $\alpha=0.001$, with $\beta_1=0.5$ and $\beta_2=0.9$.
% Its inner first and second moment vectors are reset at
% each outer epoch in order to avoid building momentum in staled directions.
For
estimating $\lambda^*$, we parametrize $\bftheta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
$q(\bftheta|\bfpsi)$ initialized with a mean at $\log(5)$ and unit variance. At
each iteration, parameters $\bfpsi$ are updated by taking one Adam step and the same hyper-parameters as for the critic.
The gradient penalty coefficient is set to
$\lambda=0.001$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=3$.

@@

As a second illustrative example, we consider a generator producing
5-dimensional continuous data, as originally specified in Section 4.2 of
\citep{cranmer2015approximating}.
More specifically, we consider the following
generative process:
\begin{itemize}
\item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
$z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
$z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
$z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
$z_3 \sim {\text{Exponential}(\lambda=3)}$, and
$z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
\item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
semi-positive definite $5 \times 5$ matrix defining a fixed projection
of $\bfz$ into the observed space.
\end{itemize}
Again, AVO does not have access to the density or its gradient, only samples from the generative model.
We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored  (mean field) Gaussian
proposal distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=50$, $n_\text{critic}=100$, $\lambda=0.01$, $\gamma=1$ and
Adam configured with a step size $\alpha=0.001$ with $\beta_1=0.5$ and $\beta_2=0.9$.
The architecture of the critic is the same as in the previous example.

@@

In this example, we consider observed data generated with the simplified generator of Eqn.~\ref{eq:simple_weinberg}
using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=1.1)$. The simulator parameters are modeled with a
factored (mean field) Gaussian proposal distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=50$,
$n_\text{critic}=100$, $\lambda=0.0001$, $\gamma=5$ and Adam configured with $\alpha=0.001$,
$\beta_1=0.5$ and $\beta_2=0.9$.
% As with the first example, we compare
% entropy penalties $\gamma=0$ and $\gamma=5$. \glnote{Double check}

The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
proposal distributions $q(\bftheta|\bfpsi)$ after
AVO. We see that the proposal distribution concentrates
in the neighborhood of $\bftheta^*$.  Despite the width of the resulting proposal distribution with respect to $\bftheta^*$
in the top right plot of  Figure~\ref{fig:weinberg}. This suggests either a relatively flat
landscape around $\bftheta^*$ or that the observed data can in this case also be
reproduced with the  predictive distribution $\qxpsi$.
Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
the variational upper bound $-U_d$ fall
towards $0$, which indicates  convergence towards a distribution that the critic
cannot distinguish from the true distribution.

>>>>

\begin{keywords}
Likelihood-free inference, implicit models, adversarial training, variational inference.
\end{keywords}

@@

As explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} (WGANs) reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, such that
\begin{align}
& W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) \nonumber \\
&= \sup_{\bfphi}  \mathbb{E}_{\tilde{\mathbf{x}}\sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}}; \bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x}; \bfphi)] \nonumber \\
&\triangleq \sup_{\bfphi} \mathcal{L}_W.
\end{align}
Under the WGAN-GP formulation of \citep{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\bfphi$ and $\bftheta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,&  \mathcal{L}_W + \lambda_{GP} \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - \mathcal{L}_W
\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

The observed data is sampled from a Poisson with mean $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 3000 iterations with mini-batches of size $M=50$
and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=100$ iterations with a step size $\alpha=0.001$, with $\beta_1=0.5$ and $\beta_2=0.9$.
% Its inner first and second moment vectors are reset at
% each outer epoch in order to avoid building momentum in staled directions.
For
estimating $\lambda^*$, we parametrize $\bftheta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
$q(\bftheta|\bfpsi)$ initialized with a mean at $\log(5)$ and unit variance. At
each iteration, parameters $\bfpsi$ are updated by taking one Adam step and the same hyper-parameters as for the critic.
The gradient penalty coefficient is set to
$\lambda_{GP}=0.001$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=3$.

@@

As a second illustrative example, we consider a generator producing
5-dimensional continuous data, as originally specified in Section 4.2 of
\citep{cranmer2015approximating}.
More specifically, we consider the following
generative process:
\begin{itemize}
\item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
$z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
$z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
$z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
$z_3 \sim {\text{Exponential}(\lambda=3)}$, and
$z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
\item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
semi-positive definite $5 \times 5$ matrix defining a fixed projection
of $\bfz$ into the observed space.
\end{itemize}
Again, AVO does not have access to the density or its gradient, only samples from the generative model.
We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored  (mean field) Gaussian
proposal distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=50$, $n_\text{critic}=100$, $\lambda_{GP}=0.01$, $\gamma=1$ and
Adam configured with a step size $\alpha=0.001$ with $\beta_1=0.5$ and $\beta_2=0.9$.
The architecture of the critic is the same as in the previous example.

@@

In this example, we consider observed data generated with the simplified generator of Eqn.~\ref{eq:simple_weinberg}
using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=1.1)$. The simulator parameters are modeled with a
factored (mean field) Gaussian proposal distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=50$,
$n_\text{critic}=100$, $\lambda=0.0001$, $\gamma=5$ and Adam configured with $\alpha=0.001$,
$\beta_1=0.5$ and $\beta_2=0.9$. For the critic $d$, we use a 3-layer MLP with 50
hidden nodes per layer and ReLU activations.
% As with the first example, we compare
% entropy penalties $\gamma=0$ and $\gamma=5$. \glnote{Double check}

The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
proposal distributions $q(\bftheta|\bfpsi)$ after
AVO. We see that the proposal distribution concentrates
in the neighborhood of $\bftheta^*$.
Despite the width of the resulting proposal distribution with respect to $\bftheta^*$,
the synthetic data closely match with the observed data, as shown
in the top right plot of  Figure~\ref{fig:weinberg}. This suggests either a relatively flat
landscape around $\bftheta^*$ or that the observed data can in this case also be
reproduced with the  predictive distribution $\qxpsi$.
Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
the variational upper bound $-U_d$ fall
towards $0$, which indicates  convergence towards a distribution that the critic
cannot distinguish from the true distribution.
